---
title: "Central statistical models and longitudinal contrasts"
author: "dar authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Central statistical models and longitudinal contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dar)
```

`add_model()` defines one statistical design shared by every compatible
differential-abundance step. Variables stored in `var_info` remain the targets
of inference; other fixed terms adjust for confounding. Method-specific
thresholds, normalization and rarefaction settings remain active.

## Cross-sectional adjustment

```{r cross-sectional, eval=FALSE}
rec <- recipe(phy, var_info = "condition", tax_info = "Species") |>
  add_model(
    ~ condition + batch + age,
    reference = c(condition = "control"),
    na_action = "error"
  ) |>
  step_deseq() |>
  step_aldex() |>
  step_ancom()

get_model(rec)
fit <- prep(rec)
```

The reference is the denominator, so a positive effect for
`condition[treated-control]` means higher abundance in `treated`. With
`na_action = "drop"`, `prep()` removes one common complete-case cohort before
any engine runs and records the sample identifiers in the execution manifest.

## Longitudinal adjustment

Numeric time values are converted internally to a factor whose levels are
sorted by numeric value. The original sample metadata are not modified.

```{r longitudinal, eval=FALSE}
longitudinal <- recipe(
  longitudinal_phy,
  var_info = "condition",
  tax_info = "Species"
) |>
  add_model(
    ~ condition * visit + batch + (1 | subject),
    time = "visit",
    reference = c(condition = "control", visit = "0"),
    contrasts = "all",
    na_action = "drop"
  ) |>
  step_ancom() |>
  step_maaslin()

fit <- prep(longitudinal)
fit@execution$contrasts
```

For `condition * visit`, the simple plan contains condition comparisons within
every visit and all pairwise visit comparisons within every condition. Later
time values are numerators. `contrasts = "all"` additionally requests
difference-in-differences. More than 12 time points requires an explicit larger
`max_time_levels`, preventing accidental contrast explosions.

## Engine capabilities

| Engine | Fixed confounders | Time interaction | Random effects |
|:--|:--:|:--:|:--:|
| DESeq2 | yes | yes | no |
| ALDEx2 | yes | yes | no |
| ANCOM-BC2 | yes | yes | yes |
| corncob | yes | yes | no |
| MaAsLin3 | yes | yes | yes |
| Wilcoxon | no | no | no |
| LEfSe | no | no | no |

Incompatible steps are warned about when the chain is built and rechecked
after preprocessing. `prep()` records them rather than treating them as failed
fits. If no method is compatible, preparation stops. Real engine failures and
missing planned contrasts remain errors.

Prepared modeled results contain `contrast_id`, `comparison`, `contrast_type`
and `var`. Consensus operations group by `contrast_id` and effect direction, so
different longitudinal hypotheses cannot contribute to the same vote.

`export_steps()` and `import_steps()` preserve the formula, time column,
references, contrast mode, missing-value policy and time-level limit. Files
created by older versions without a model continue to import through the
legacy path.
