---
title: "Central statistical models and longitudinal contrasts"
author: "dar authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Central statistical models and longitudinal contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dar)
```

`add_model()` defines one statistical design shared by every compatible
differential-abundance step. Its `targets` and `tax_level` arguments are the
single source of truth for the analysis configuration; other fixed terms adjust
for confounding. Method-specific thresholds, normalization and rarefaction
settings remain active. A model-free recipe remains valid for preprocessing,
but differential-abundance execution without a model is deprecated.

## Cross-sectional adjustment

```{r cross-sectional, eval=FALSE}
rec <- recipe(phy) |>
  add_model(
    ~ condition + batch + age,
    targets = "condition",
    tax_level = "Species",
    reference = c(condition = "control"),
    na_action = "error"
  ) |>
  step_deseq() |>
  step_aldex() |>
  step_ancom()

get_model(rec)
fit <- prep(rec)
```

The reference is the denominator, so a positive effect for
`condition[treated-control]` means higher abundance in `treated`. With
`na_action = "drop"`, `prep()` removes one common complete-case cohort before
any engine runs and records the sample identifiers in the execution manifest.

## Longitudinal adjustment

Numeric time values are converted internally to a factor whose levels are
sorted by numeric value. The original sample metadata are not modified.

```{r longitudinal, eval=FALSE}
longitudinal <- recipe(longitudinal_phy) |>
  add_model(
    ~ condition * visit + batch + (1 | subject),
    targets = "condition",
    tax_level = "Species",
    time = "visit",
    reference = c(condition = "control", visit = "0"),
    contrasts = "all",
    na_action = "drop"
  ) |>
  step_ancom() |>
  step_maaslin()

fit <- prep(longitudinal)
fit@execution$contrasts
```

For `condition * visit`, the simple plan contains condition comparisons within
every visit and all pairwise visit comparisons within every condition. Later
time values are numerators. `contrasts = "all"` additionally requests
difference-in-differences. More than 12 time points requires an explicit larger
`max_time_levels`, preventing accidental contrast explosions.

## Engine capabilities

| Engine | Fixed confounders | Time interaction | Random effects |
|:--|:--:|:--:|:--:|
| DESeq2 | yes | yes | no |
| ALDEx2 | yes | yes | no |
| ANCOM-BC2 | yes | yes | yes |
| corncob | yes | yes | no |
| MaAsLin3 | yes | yes | yes |
| Wilcoxon | no | no | no |
| LEfSe | no | no | no |

Incompatible steps are warned about when the chain is built and rechecked
after preprocessing. `prep()` records them rather than treating them as failed
fits. If no method is compatible, preparation stops. Real engine failures and
missing planned contrasts remain errors.

Prepared modeled results contain `contrast_id`, `comparison`, `contrast_type`
and `var`. Consensus operations group by `contrast_id` and effect direction, so
different longitudinal hypotheses cannot contribute to the same vote.

`export_steps()` and `import_steps()` preserve the formula, targets, taxonomic
level, time column, references, contrast mode, missing-value policy and
time-level limit. Files created by older versions without a model continue to
import through the legacy path.

## Migrating legacy selectors

The former `recipe(..., var_info=, tax_info=)`, `add_var()`, `add_tax()`,
`get_var()` and `get_tax()` interfaces remain available during the first
Bioconductor deprecation cycle. They emit classified warnings and synchronize
with the centralized model where applicable. New code should inspect
`get_model(rec)$targets` and `get_model(rec)$tax_level` instead. Legacy recipes
can be migrated by moving both selectors into `add_model()`:

```{r migration, eval=FALSE}
# Deprecated
old <- recipe(phy, var_info = "condition", tax_info = "Species")

# Current API
current <- recipe(phy) |>
  add_model(
    ~ condition,
    targets = "condition",
    tax_level = "Species"
  )
```
