---
title: "Central statistical models and longitudinal contrasts"
author: "dar authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Central statistical models and longitudinal contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dar)
```

`add_model()` defines one statistical design shared by every compatible
differential-abundance step. Its `targets` and `tax_level` arguments are the
single source of truth for the analysis configuration; other fixed terms adjust
for confounding. Method-specific thresholds, normalization and rarefaction
settings remain active. A model-free recipe remains valid for preprocessing,
but differential-abundance execution without a model is deprecated.

## Cross-sectional adjustment

```{r cross-sectional, eval=FALSE}
rec <- recipe(phy) |>
  add_model(
    ~ condition + batch + age,
    targets = "condition",
    tax_level = "Species",
    reference = c(condition = "control"),
    na_action = "error"
  ) |>
  step_deseq() |>
  step_aldex() |>
  step_ancom()

get_model(rec)
fit <- prep(rec)
```

The reference is the denominator, so a positive effect for
`condition[treated-control]` means higher abundance in `treated`. With
`na_action = "drop"`, `prep()` removes one common complete-case cohort before
any engine runs and records the sample identifiers in the execution manifest.

## Longitudinal adjustment

Numeric time values are converted internally to a factor whose levels are
sorted by numeric value. The original sample metadata are not modified.

```{r longitudinal, eval=FALSE}
longitudinal <- recipe(longitudinal_phy) |>
  add_model(
    ~ condition * visit + batch + (1 | subject),
    targets = "condition",
    tax_level = "Species",
    time = "visit",
    reference = c(condition = "control", visit = "0"),
    contrasts = "all",
    na_action = "drop"
  ) |>
  step_ancom() |>
  step_maaslin()

fit <- prep(longitudinal)
fit@execution$contrasts
```

For `condition * visit`, the simple plan contains condition comparisons within
every visit and all pairwise visit comparisons within every condition. Later
time values are numerators. `contrasts = "all"` additionally requests
difference-in-differences. More than 12 time points requires an explicit larger
`max_time_levels`, preventing accidental contrast explosions.

## Engine capabilities

| Engine | Fixed confounders | Time interaction | Random effects |
|:--|:--:|:--:|:--:|
| DESeq2 | yes | yes | no |
| ALDEx2 | yes | yes | no |
| ANCOM-BC2 | yes | yes | yes |
| corncob | yes | yes | no |
| MaAsLin3 | yes | yes | yes |
| Wilcoxon | no | no | no |
| LEfSe | no | no | no |

Incompatible steps are warned about when the chain is built and rechecked
after preprocessing. `prep()` records them rather than treating them as failed
fits. If no method is compatible, preparation stops. Real engine failures and
missing planned contrasts remain errors.

Prepared modeled results contain `contrast_id`, `comparison`, `contrast_type`
and `var`. Consensus operations group by `contrast_id` and effect direction, so
different longitudinal hypotheses cannot contribute to the same vote.

## A uniform tidy result contract

`tidy_results()` exposes the same columns for every DA engine while preserving
the engine-specific tables in `fit@results`. The `method` column identifies the
stable engine (`deseq`, `aldex`, and so on), whereas `step_id` identifies the
configured instance and therefore distinguishes repeated configurations of one
engine.

```{r tidy-results, eval=FALSE}
results <- tidy_results(
  fit,
  steps = c("ancom__primary", "maaslin__sensitivity"),
  significant_only = TRUE
)

results |>
  dplyr::select(
    taxa_id, contrast_id, comparison, effect_size, effect_metric,
    adj_p_value, significant, method, step_id
  )
```

Effect values retain their native scientific scale. Use `effect_metric` when
interpreting them: for example, a DESeq2 log2 fold change and an ALDEx2
standardized CLR effect are labelled explicitly but are not converted to a
common scale and should not be compared numerically.

## Advanced native-engine arguments

The seven DA steps accept `engine_args`, a named list grouped by native-call
stage. This keeps advanced settings attached to the configured step and avoids
an ambiguous `...`. For example:

```{r engine-args, eval=FALSE}
rec |>
  step_deseq(
    engine_args = list(
      fit = list(minReplicatesForReplace = Inf),
      shrink = list(lfcThreshold = 1)
    )
  ) |>
  step_ancom(
    engine_args = list(
      fit = list(iter_control = list(tol = 1e-3, max_iter = 50))
    )
  )
```

Available stages are `size_factors`, `fit`, `results`, and `shrink` for
DESeq2; `clr`, `fit`, and `effect` for ALDEx2; `fit` for ANCOM-BC2 and
corncob; `fit` and `contrast` for MaAsLin3; `test` for Wilcoxon; and
`transform` and `fit` for LEfSe. Data, formulas, contrasts, output paths and
arguments already exposed by each dar wrapper are protected. Non-empty
`engine_args` require a centralized model. A configured stage that is not used
by the selected execution route is reported as an error.

`export_steps()` and `import_steps()` preserve the formula, targets, taxonomic
level, time column, references, contrast mode, missing-value policy and
time-level limit. Files created by older versions without a model continue to
import through the legacy path.

## Migrating legacy selectors

The former `recipe(..., var_info=, tax_info=)`, `add_var()`, `add_tax()`,
`get_var()` and `get_tax()` interfaces remain available during the first
Bioconductor deprecation cycle. They emit classified warnings and synchronize
with the centralized model where applicable. New code should inspect
`get_model(rec)$targets` and `get_model(rec)$tax_level` instead. Legacy recipes
can be migrated by moving both selectors into `add_model()`:

```{r migration, eval=FALSE}
# Deprecated
old <- recipe(phy, var_info = "condition", tax_info = "Species")

# Current API
current <- recipe(phy) |>
  add_model(
    ~ condition,
    targets = "condition",
    tax_level = "Species"
  )
```

```{r session-info}
sessionInfo()
```
