---
title: "dar scientific validation"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M UTC', tz = 'UTC')`"
output:
  html_document:
    toc: true
    toc_float: true
params:
  data_dir: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
read_artifact <- function(name) {
  path <- file.path(params$data_dir, paste0(name, ".csv"))
  if (!file.exists(path) || file.info(path)$size == 0L) return(data.frame())
  tryCatch(utils::read.csv(path, stringsAsFactors = FALSE), error = function(cnd) data.frame())
}
runs <- read_artifact("runs")
manifests <- read_artifact("manifests")
summary <- read_artifact("summary")
gates <- read_artifact("gates")
```

## Scope

This report characterizes repeated-sampling behavior of the centralized differential-abundance models. It is development evidence, not a ranking of engines. P-values and effect magnitudes are never compared directly across methods with different estimands.

## Simulation manifest

```{r manifest}
if (nrow(manifests)) {
  manifest_summary <- manifests |>
    dplyr::group_by(.data$scenario) |>
    dplyr::summarise(
      replicates = dplyr::n_distinct(.data$replicate),
      samples_min = min(.data$samples),
      samples_max = max(.data$samples),
      taxa = max(.data$taxa),
      contrasts = max(.data$contrasts),
      .groups = "drop"
    )
  knitr::kable(manifest_summary)
} else {
  cat("No simulation manifests were available.")
}
```

## Engine execution

```{r execution}
if (nrow(runs)) {
  execution <- runs |>
    dplyr::count(.data$engine, .data$scenario, .data$status, name = "runs")
  knitr::kable(execution)
} else {
  cat("No engine runs were available.")
}
```

## Statistical gates

```{r gates-table}
if (nrow(gates)) {
  gate_counts <- gates |>
    dplyr::count(.data$engine, .data$status, name = "gates")
  knitr::kable(gate_counts)
} else {
  cat("No statistical gates were evaluated.")
}
```

```{r failed-gates}
if (nrow(gates) && any(gates$status == "fail")) {
  failed <- gates |>
    dplyr::filter(.data$status == "fail") |>
    dplyr::select(
      .data$engine, .data$scenario, .data$contrast_id, .data$metric,
      .data$estimate, .data$lower, .data$upper, .data$limit, .data$n_replicates
    )
  knitr::kable(failed)
} else {
  cat("No evaluated gate failed.")
}
```

## Metric estimates

```{r metric-plot, fig.width=11, fig.height=8}
if (nrow(summary)) {
  plotted <- summary |>
    dplyr::filter(.data$metric %in% c(
      "type1_error", "fdr", "power", "direction_recovery",
      "finite_rate", "execution_rate", "coverage"
    ))
  if (nrow(plotted)) {
    ggplot2::ggplot(
      plotted,
      ggplot2::aes(x = .data$engine, y = .data$estimate, colour = .data$scenario)
    ) +
      ggplot2::geom_point(position = ggplot2::position_dodge(width = 0.5)) +
      ggplot2::geom_errorbar(
        ggplot2::aes(ymin = .data$lower, ymax = .data$upper),
        width = 0.2,
        position = ggplot2::position_dodge(width = 0.5)
      ) +
      ggplot2::facet_wrap(~ metric, scales = "free_y") +
      ggplot2::coord_cartesian(ylim = c(0, 1)) +
      ggplot2::theme_minimal() +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
  }
}
```

## Interpretation limits

- Monte Carlo intervals quantify uncertainty between independent simulation replicates.
- Bias and coverage are evaluated only when the reported effect and simulation truth share a documented scale.
- `not_applicable` is intentional for incompatible estimands or unavailable uncertainty estimates.
- `informational` means the selected profile did not reach the minimum replicate count for a regression gate.
- Expected capability skips are not engine failures; errors from compatible models are failures.
